Implementation and assessment of tools for integrating forest roads into projections of the cumulative effects of disturbance on wildlife

### 2024-05-08 analysis/paper/paper.Rmd
writing in progress

Despite a rapidly growing body of evidence of complex and varied impacts of road networks on wildlife and other values, and increasing use of spatial stochastic models for forecasting impacts of forest disturbance across regions, very few spatial stochastic projections of forest disturbance dynamics have included road network projections, in part because tools suited for this purpose have not been available. Hardy et al. [-@hardy_forest_2019; -@hardy_landis-ii_2023] partially addressed this need with an extension that can be integrated with other components of the Landis-II modelling framework [e.g. @labadie_global_2023]. Our implementations of simple iterative least cost path and minimum spanning tree methods [@lochhead_demo_2018; @lochhead_linking_2022] in the open source R package `roads` [@endicott_roads_2023] provide a complementary and comparable alternative to Hardy's method that can be easily integrated into R workflows and modelling frameworks used for forecasting forest change in Canada and elsewhere [e.g. @barros_empowering_2023; @daniel_state-and-transition_2016].

We evaluated performance of these algorithms in a topographically complex region of British Columbia where we anticipated projection of forest road network patterns would be challenging, using a very simple cost surface model. Thus, we were not surprised to see less accurate projections of road density and presence than in less mountainous regions of eastern Canada with more careful accounting for the cost of water crossings [@hardy_landis-ii_2023]. Water crossing behaviour in Revelstoke could be improved with a more detailed high resolution cost surface, but none of these simple algorithms will project switchbacks on steep slopes that are characteristic of forest roads in mountains. A related fundamental limitation is that none of these algorithms can project road tortuosity within raster cells, so in complex landscapes projections of road density become less accurate as raster cell resolution increases.

Aspatial metrics of overall projection accuracy can be difficult to interpret. To more deeply understand variation in the behaviour of road network projection algorithms we found it helpful to distinguish between performance within cutblocks and outside of them, and to examine spatial variation in discrepancies. Examination of spatial mismatch between observed and projected roads highlighted errors in our input data, and removing mapped cutblocks without observed access roads from our analysis helped resolve the apparently contradictory results. These input data errors are not unique to our study area; there is considerable variation in the consistency and accuracy of available resource road information across Canada and elsewhere [@poley_identifying_2022; @ministry_of_forests_resource_nodate] that users should keep in mind interpreting discrepancies between observed and projected road networks.

Outside cutblocks, we saw very little variation in performance among projection method variants. All were somewhat better than nothing, but the improvement was small because in this landscape new landings were fairly close to existing roads. Outside cutblocks our ILCP algorithm is similar to Hardy's algorithm, so we expect little difference in behaviour where targets are far from existing roads. Within cutblocks, the density of landing points had a much larger impact on performance than the choice of DCLP or MST algorthm. The Hardy algorithm most accurately projected high road density within blocks, but examination of spatial patterns clarified that the projected tortuousity is quite different from observed switchbacks, and may be an artefact of searching for nearest destinations \>200 m away on a 100 m grid. One could increase the density of landing points in the DCLP/MST projections to match the observed high density of roads within cutblocks, but if realistic switchbacks in mountainous terrain is a key goal then none of these algorithms are sufficient.

Simple algorithms can recreate some aspects of observed road networks but not others, and more complex realistic algorithms are more costly to develop and use. Before investing in additional complexity, it is important to recognize that adequacy depends on the goal. For some purposes, projections from relatively crude cost models and simple algorithms may be adequate. For example, models of caribou movement often include distance to roads as a predictor [e.g. @labadie_global_2023], and a combined metric of buffered anthropogenic disturbance best explains variation in boreal caribou demographic rates among ranges [@johnson_science_2020]. In our example landscape, projections of size disturbance footprints (500 m and 1km buffer) and distance to road are reasonably accurate, and for wildlife response models informed by these measures our somewhat crude road projections are likely good enough.

>>>>

Despite a rapidly growing body of evidence of complex and varied impacts of road networks on wildlife and other values, and increasing use of spatial stochastic models for forecasting impacts of forest disturbance across regions, very few spatial stochastic projections of forest disturbance dynamics have included road network projections, in part because tools suited for this purpose have not been available. Hardy et al. [-@hardy_forest_2019; -@hardy_landis-ii_2023] partially addressed this need with an extension that can be integrated with other components of the Landis-II modelling framework [e.g. @labadie_global_2023]. Our implementations of iterative least cost path and minimum spanning tree methods [@lochhead_demo_2018; @lochhead_linking_2022] in the open source R package `roads` [@endicott_roads_2023] provide an alternative to Hardy's method that better predicts road patterns in our example landscape, and can be easily integrated into R workflows and modelling frameworks used for forecasting forest change in Canada and elsewhere [e.g. @barros_empowering_2023; @daniel_state-and-transition_2016]. Although projecting roads within ecological forecasting models was beyond the scope of this study, our analysis of performance can assist with method selection in this context.

We evaluated performance of forest road network projection algorithms in a topographically complex region of British Columbia where we anticipated projection would be challenging, and implemented a simplified version of Anderson and Nelson’s (2004) grade penalty method for setting graph edge weights that improves performance in this mountainous region. The simplified grade penalty method does not increase memory or computational requirements, and requires only a few additional lines of code to implement (Supplement A). We see no disadvantages to the approach in regions where slope is an important determinant of road construction cost, and recommend using it. However, we also acknowledge that the important determinants of road construction costs may vary among landscapes, and have also allowed the possibility for users of the `roads` R package to define their own method for determining edge weights from adjacent node values.

Among the method variants we considered, the minimum spanning tree (MST) regular high density method yielded better projections of average road density and the forestry disturbance footprint (Figure 3). Aspatial projections of road presence, distance to road, and road disturbance footprint were comparable to Hardy's method [-@hardy_landis-ii_2023] (Figure 3), so overall the MST regular high density projections were more accurate. However, the MST algorithm is also more computationally demanding than the other methods we considered (Supplement E), so users will need to consider whether the benefits of improved performance outweigh the increased computational costs in their particular circumstance.

Aspatial metrics of overall projection accuracy can be difficult to interpret. To more deeply understand variation in the behaviour of road network projection algorithms we found it helpful to distinguish between performance within cutblocks and outside of them, and to examine spatial patterns. This examination highlighted some errors that could be fixed by better accounting for the cost of water crossing in our weight raster, and other more fundamental limitations and differences among methods. Even in cases where the Hardy and grade penalty methods project similar mean road densities, the projected pattern of roads within cutblocks is quite different, in that the grade penalty roads tend to follow elevation contours, while the Hardy method produces curves (Figure 4). We suspect the latter behaviour is an artifact of searching for nearest destinations \>200 m away on a 100 m grid.

Although the grade penalty method improves performance by aligning roads with elevation contours, all these methods create dead ends, intersections, and abrupt turns that are avoided in real road networks. We chose not to implement Anderson and Nelson's methods (2004) for penalizing sharp curves and large junction angles because these complications would increase algorithmic complexity and computational costs. Simple algorithms can recreate some aspects of observed road networks but not others, and more complex realistic algorithms are generally more costly to develop and use. Before investing in additional complexity, it is important to recognize that adequacy depends on the goal. For some purposes, projections from relatively crude cost models and simple algorithms may be adequate. For example, models of caribou movement often include distance to roads as a predictor [e.g. @labadie_global_2023], and a combined metric of buffered anthropogenic disturbance best explains variation in boreal caribou demographic rates among ranges [@johnson_science_2020]. In our example landscape, projections of disturbance footprints (500 m and 1km buffer) and distance to road are reasonably accurate, and for wildlife response models informed by these measures our somewhat crude road projections are likely good enough.

Examination of spatial mismatch between observed and projected roads also highlighted errors in our input data. We note these errors because they are not unique to our study area; there is considerable variation in the consistency and accuracy of available resource road information across Canada and elsewhere [@poley_identifying_2022; @ministry_of_forests_resource_nodate] that users should keep in mind interpreting discrepancies between observed and projected road networks.

These road projection methods require construction of large graphs, with a node for each pixel in which road construction is possible. The associated memory and computational requirements may be prohibitive for large landscapes (Supplement E). Options for reducing those requirements include restricting the area where road construction is possible (by setting some portions of the weight raster to NA), reducing landing density within cutblocks (by using the centroid method), increasing raster resolution, or using Hardy et al.'s less costly implementations [-@hardy_forest_2019; -@hardy_landis-ii_2023]. The first three options are available to users of the `roads` R package. However, those contemplating reducing computational cost by increasing raster resolution should keep in mind that within grid cells all of these methods project straight roads, so in complex landscapes projections of road density become less accurate as raster cell resolution increases (Supplement D). The distance to road metric is less sensitive to changes in cell resolution, and distance metrics may also be more ecologically meaningful [@rowland_elk_2000]. More labour intensive possibilities for reducing computational costs include implementing slow components of the R package (Supplement F) in a faster compiled language, and implementing a multi-scale approach in which coarse resolution results are used to limit the area considered in finer resolution projections.

### 2024-05-08 analysis/paper/supplement.Rmd
writing in progress

slopePenaltyFn <- function(x1, x2, resolution = 1, baseCost = 16178, limit = 20,
                           penalty = 504, limitWeight = NA){
  # If one of the nodes is a road or barrier ignore grade penalty
  cond <- pmin(x1, x2) >= 0
  cond[is.na(cond)] <- F

  # percent slope, if both nodes have elevation values.
  grade <- 100 * abs(x1 - x2) * cond / resolution

@@

  # if both 0 this is an existing road link. Otherwise it is a barrier.
  slp[!cond] <- abs(pmin(x1, x2))[!cond]

@@

```{r profiling-ilcp, fig.cap="Profiling results for the ILCP method showing functions that take > 1 s to run on a cost raster with 1 million cells and 10 landings. Functions higher in the call stack are called by functions lower down.", fig.height=2}
if(!file.exists(here::here("analysis/data/derived_data/ilcp_Rprof.out"))){
  source(here::here("analysis/scripts/test_profile.R"), keep.source = TRUE)
}

profstat(here::here("analysis/data/derived_data/ilcp_Rprof.out"))

@@

```{r profiling-mst, fig.cap="Profiling results for the MST method showing functions that take > 1 s to run on a cost raster with 1 million cells. Functions higher in the call stack are called by functions lower down.", fig.height=5}
profstat(here::here("analysis/data/derived_data/mst_Rprof.out"))
```

>>>>

slopePenaltyFn <- function(x1, x2, resolution = 1, baseCost = 16178, limit = 20,
                           penalty = 504, limitWeight = NA){
  # Don't calculate grade penalty cost if one of the nodes is a barrier.
  cond <- pmin(x1, x2) >= 0
  cond[is.na(cond)] <- F

  # Apply grade penalty if both nodes have elevation values. 
  # If one node is a road use base cost.
  grade <- 100 * abs(x1 - x2) * cond / resolution

@@

  # If both 0 this is an existing road link. Otherwise it is a barrier.
  slp[!cond] <- abs(pmin(x1, x2))[!cond]

@@

```{r profiling-ilcp, fig.cap="Profiling results for the ILCP method showing functions that take > 1 s to run on a cost raster with 1 million cells and 10 landings. Functions higher in the call stack are called by functions lower down.", fig.height=2}
if(!file.exists(here::here("analysis/data/derived_data/ilcp_100_Rprof.out"))){
  source(here::here("analysis/scripts/test_profile.R"), keep.source = TRUE)
}

profstat(here::here("analysis/data/derived_data/ilcp_100_Rprof.out"))

@@

```{r profiling-mst, fig.cap="Profiling results for the MST method showing functions that take > 1 s to run on a cost raster with 1 million cells. Functions higher in the call stack are called by functions lower down.", fig.height=5}
profstat(here::here("analysis/data/derived_data/mst_100_Rprof.out"))
```
